### 2025-06-01 conteudo/background/linux.tex
fix grammar with grammarly

This chapter presents an overview of fundamental topics needed to understand 
the research, such as Linux Kernel, refactoring, types of code duplication, 
code duplication detection methods presented in the literature and ways to 
evaluating these methods. This foundation will make it easier to grasp the 
concepts discussed in the research.

@@

The Linux kernel is monolithic and organized into a set of subsystems, such as
the process scheduler, memory management, device driver infrastructure,
networking, and filesystems \citep{melissa}. Each subsystem typically has a
dedicated maintainer or a team of maintainers responsible for overseeing its
development and managing contributions \citep{melissa}.

Development contributions to the Linux kernel are coordinated using the Git
version control system. These contributions are formatted as patches—text
documents that outline the differences between two versions of the source code.
These patches are then submitted and reviewed via the Linux mailing lists
\citep{melissa}.

@@

One of the subsystems within the Linux kernel is the AMD Display driver, which
is responsible for implementing the drivers needed for AMD GPUs to function
correctly in a Linux environment. The maintainers of this subsystem have
reported a practical issue: a significant amount of duplicated code, which
complicates the maintenance of this subsystem. This problem serves as the
starting point for our research, with the AMD Display Driver as our primary
object of study as we aim to understand and mitigate code duplication in the
Linux kernel. The AMD Display driver subsystem contains more then 391000 lines 
of code and more then 1089 code files
\footnote{ 
Data extracted directly from the Linux repository with the help of the cloc tool.
\citep{cloc}
}, 
making it a considerable component within the Linux kernel. Like other 
subsystems, discussions about the AMD Display driver are conducted on its 
designated mailing list, which can be accessed here: 
\url{https://lists.freedesktop.org/mailman/listinfo/amd-gfx}.

Another subsytem within the Linux kernel explored in this work is the Industrial
I/O subsytem (IIO). From the Linux kernel documentation, ''The main purpose of the 
Industrial I/O subsystem (IIO) is to provide support for devices that in some sense 
perform either analog-to-digital conversion (ADC) or digital-to-analog conversion (DAC) or both``\citep{iiodoc}.
The IIO subsystem contains more then X lines of code and more then Y code files 
\footnote{
Data extracted directly from the Linux repository with the help of the clol tool. \citep{cloc}
}.
We opted to include the IIO subsystem in our experiments to ensure our findings are not limited to a single subsystem. 
Additionally, we have an established point of contact within the IIO subsystem to facilitate this work.
. Like other subsystems, discussions about the IIO subsytem are conducted on its 
designated mailing list, which can be found here: 
\url{https://subspace.kernel.org/vger.kernel.org.html}.

@@

The \textit{diff} command is a command line in Linux that allows the user to compare 
two files line by line \citep{diffcommand}. There are multiple forms 
that the results of the command can be shown, one example of the command being executed 
can be found in Figure \ref{fig:diff}. With the command results, it is easy to visualize and extract 
the number of lines that are equal and different between the two compared files.

\begin{figure}
\includegraphics[scale=0.3]{diff_command}
\caption{Example of \textit{diff} command executed}
File 1 content is shown in red and file 2 content is shown in green. \textit{-} 
represents a line that only exist in file 1, \textit{+} represents a line that only exist in file 2,
\textit{!} represents a line that exist in both file but has modifications between them, and empty on the
start of a line represents that the line exist in both files without any modifications.
\label{fig:diff}
\end{figure}

We use thhe \textit{diff} command as the base of one of the code duplication detection methods 
on the tool proposed on this work, although it is not extensive explored as the text similarity
method for code duplication detection.

>>>>

This chapter presents an overview of fundamental topics needed to understand 
the research, such as the Linux Kernel, refactoring, types of code duplication, 
code duplication detection methods presented in the literature and ways to 
evaluate these methods. This foundation will make it easier to grasp the 
concepts discussed in the research.

@@

The Linux kernel is monolithic and organized into subsystems, such as
the process scheduler, memory management, device driver infrastructure,
networking, and filesystems \citep{melissa}. Each subsystem typically has a
dedicated maintainer or a team of maintainers responsible for overseeing its
development and managing contributions \citep{melissa}.

Development contributions to the Linux kernel are coordinated using the Git
version control system. These contributions are formatted as patches—text
documents that outline the differences between two source code versions.
These patches are then submitted and reviewed via the Linux mailing lists
\citep{melissa}.

@@

One of the subsystems within the Linux kernel is the AMD Display driver, which
is responsible for implementing the drivers needed for AMD GPUs to function
correctly in a Linux environment. The maintainers of this subsystem have
reported a practical issue: a significant amount of duplicated code, which
complicates the maintenance of this subsystem. This problem serves as the
starting point for our research, with the AMD Display driver as our primary
object of study as we aim to understand and mitigate code duplication in the
Linux kernel. The AMD Display driver subsystem contains more then 391000 lines 
of code and more then 1089 code files
\footnote{ 
Data extracted directly from the Linux repository with the help of the cloc tool.
\citep{cloc}
}, 
making it a considerable component within the Linux kernel. Like other 
subsystems, discussions about the AMD Display driver are conducted on its 
designated mailing list, which can be accessed here: 
\url{https://lists.freedesktop.org/mailman/listinfo/amd-gfx}.

Another subsystem within the Linux kernel explored in this work is the
Industrial I/O subsystem (IIO). According to the Linux kernel documentation, 
''The main purpose of the Industrial I/O subsystem (IIO) is to provide support 
for devices that in some sense perform either analog-to-digital conversion (ADC) 
or digital-to-analog conversion (DAC) or both''\citep{iiodoc}. The IIO subsystem 
contains more than X lines of code and more than Y code files
\footnote{
Data extracted directly from the Linux repository with the help of the clol tool. \citep{cloc}
}.
We opted to include the IIO subsystem in our experiments to ensure our findings 
are not limited to one subsystem. Additionally, we have an established point of
contact within the IIO subsystem to facilitate this work.
Like other subsystems, discussions about the IIO subsytem are conducted on its 
designated mailing list, which can be found here: 
\url{https://subspace.kernel.org/vger.kernel.org.html}.

@@

The \textit{diff} command is a command line in Linux that allows the user to compare 
two files line by line \citep{diffcommand}. There are multiple forms in which the command 
results can be shown to the user, and Figure \ref{fig:diff} shows one of them. 
With the command results, it is easy to visualize and extract the number of lines that 
are equal and different between the two compared files.

\begin{figure}
\includegraphics[scale=0.25]{diff_command}
\caption{Example of \textit{diff} command executed}
File 1's content is shown in red, and file 2's content is shown in green. \textit{-} 
represents a line that only exists in file 1, \textit{+} represents a line that only 
exists in file 2,\textit{!} represents a line that exists in both files but has modifications 
between them, and empty at the start of a line represents that the line exists in 
both files without any modifications.
\label{fig:diff}
\end{figure}

We use the \textit{diff} command as the base of one of the code duplication detection 
methods on the tool proposed in this work, although it is not explored as the text 
similarity method for code duplication detection.

### 2025-06-01 conteudo/introduction/intro.tex
fix grammar with grammarly

Our studies and observations reveal that the code duplication problem described
in the previous section cannot be fully addressed by existing tools in
either the formal or grey literature. To fill this gap, we propose an approach
to identify and mitigate code duplications specifically within the Linux
kernel subsystems, focuding our study on the AMD Display Driver and the Industril IO subsytem. 
Our approach includes the development of a tool capable of detecting
duplicated functions and establishing strategies for
addressing these duplications. We tested the approach in the context of two
kernel components, the AMD Display Driver the the Industrial IO subsystem.

@@

This research aims to identify patterns of software engineering techniques, 
such as refactoring methods commonly used to address code duplications. 
These patterns will serve as a guide for future contributors in mitigating 
code duplication issues within this context. To investigate our approach, 
we are developing a tool to identify duplicate code. We are then refactoring 
portions of this code and submitting patches to maintainers, providing evidence 
of the tool's accuracy in detecting duplicate functions. Our research design follows 
a multi-method approach, structured into two phases, each employing different research methods.

In \textbf{Phase I}, we propose, develop, and validate a tool capable of
identifying code duplications at the function level within C codebases. We
expect this tool to effectively detect code duplications in the targeted subsytems of the Linux kernel,
as preliminary investigations suggest that kernel code files often contain a
mix of duplicated and unique functions. To validate the tool, we will use a
triangulation method, comparing its results with established code duplication
database referenced in the literature, and conducting an empirical analysis on
a randomly selected set of files from the AMD Display driver, which our tool
flags as containing duplications. A detailed explanation of our validation
process is provided in Chapter \ref{cha:method}.

Upon completing the first phase, we anticipate having a validated tool for
identifying code duplications within the context of the Linux kernel subsytems studied. 
We then
proceed to the second phase, which focuses on mitigating these
duplications. In \textbf{Phase II}, our objective is to verify if we can 
practically mitigate the duplication finds by the developed tool. 
To achieve this, we have designed an ethnographic approach that enables us to
interact with the Linux development community and gather artifacts that either
support or challenge our proposed mitigation strategies. Figure
\ref{fig:reDesign} summarizes the research steps proposed in this study.

@@

This manuscript consists of five more chapters. Chapter \ref{cha:back} presents
the literature overview for code duplication detection (main definitions,
current approaches in the literature), a brief description of the Linux kernel and the components
explored in this work, 
and a review of refactoring methods used throughout this research. Chapter
\ref{cha:tool} presents ArKanjo, our proposed tool to detect code duplication, 
describing all the main components.  Chapter \ref{cha:method}
describes the research methods selected to guide our work. Chapter
\ref{cha:results} shows the results we had through our work, from the research methods to 
evaluate our tool and the ethnographic studies. Chapter \ref{cha:conclusion}udes this research.

>>>>

Our studies and observations reveal that existing tools cannot fully address the code duplication 
problem described in the previous section in the formal or grey literature. To fill this gap,
we propose an approach to identify and mitigate code duplications, specifically within the
Linux kernel subsystems, focusing our study on the AMD Display driver and the Industrial 
IO subsystem. Our approach includes the development of a tool capable of detecting duplicated 
functions and establishing strategies for addressing these duplications. We tested the approach 
in the context of two kernel components, the AMD Display driver and the Industrial IO subsystem.

@@

This research aims to identify patterns of software engineering techniques, 
such as refactoring methods commonly used to address code duplications. 
These patterns will serve as a guide for future contributors in mitigating 
code duplication issues within this context. To investigate our approach, 
we are developing a tool to identify duplicate code. We then refactor portions 
of this code and submit patches to maintainers, providing evidence of the tool’s 
accuracy in detecting duplicate functions. Our research design follows a 
multi-method approach, structured into two phases, each employing different research methods.

In \textbf{Phase I}, we propose, develop, and validate a tool capable of
identifying code duplications at the function level within C codebases.
We expect this tool to effectively detect code duplications in the targeted subsystems 
of the Linux kernel, as preliminary investigations suggest that kernel code files often 
contain a mix of duplicated and unique functions. We will use a triangulation method to 
validate the tool, comparing its results with an established code duplication database 
referenced in the literature and conducting an empirical analysis on a randomly selected 
set of files from the AMD Display driver, which our tool flags as containing duplications.
A detailed explanation of our validation process is provided in Chapter \ref{cha:method}.

Upon completing the first phase, we anticipate having a validated tool for identifying 
code duplications within the context of the Linux kernel subsystems studied. We then 
proceed to the second phase, which focuses on mitigating these duplications. 
In \textbf{Phase II}, we aim to verify if we can mitigate the duplication found by 
the developed tool. To achieve this, we have designed an ethnographic approach that 
enables us to interact with the Linux development community and gather artifacts that 
either support or challenge our proposed mitigation strategies. Figure \ref{fig:reDesign} 
summarizes the research steps proposed in this study.

@@

This manuscript consists of five more chapters. Chapter \ref{cha:back} presents
the literature overview for code duplication detection (main definitions,
current approaches in the literature), a brief description of the Linux kernel and the components
explored in this work, 
and a review of refactoring methods used throughout this research. Chapter
\ref{cha:tool} presents ArKanjo, our proposed tool to detect code duplication, 
describing all the main components.  Chapter \ref{cha:method}
describes the research methods selected to guide our work. Chapter
\ref{cha:results} shows the results we had through our work, from the research methods to 
evaluate our tool and the ethnographic studies. Chapter \ref{cha:conclusion} concludes this research.

### 2025-06-01 conteudo/methods/methods.tex
fix grammar with grammarly

We employ a multi-method research approach, using different methodologies to achieve our goals
to approach code duplication in some of the Linux subsystems. 
To validate if the ArKanjo tool is capable of finding code duplications, 
applied two independent methods and perform triangulation on the results.

The first method evaluates the proposed tool using an approach from the literature, 
comparing it against the BigCloneBench dataset \citep{bigclonebench}. 
The second method is an empirical analysis of a subset of functions within the 
AMD Display driver that our tool identified as duplicates. After completing these two 
methods, we triangulated the findings to analyze the results comprehensively. 
We chose triangulation because relying solely on the literature-based method has 
limitations and may not be entirely reliable \citep{bigfail, litreview}. As a reminder, 
all research approches were done using the text similarity method as the 
code dupication method of the tool.

Given the validation of the tool's capabilities, we proceed to undertand the viability 
to mitigate the code duplication finds by the tool. 
To underdand the viability, our first choice is to conduct an ethnographic study using 
a participant-observation methodology. 
In this study, we aim to mitigate a subset of duplicated functions by refactoring them 
while interacting with the Linux community to gather feedback on our mitigation efforts.

The ethnographic study happened in two phases. 
In the first phase, the author of this research explored the process of mitigating
a code duplication found in the AMD Display Driver and sent it as a PATCH to iteract with the
Linux community and get their feedback. On a second phase, we proposed to students on the 
course Free Software Development at University São Paulo to mitigate the code duplications found 
by the tool in the IIO subsystems and the AMD Display Driver and pass the process of send the mitigations as PATCH.

@@

This section outlines our approach for mitigating code duplication in the Linux kernel
and describes the ethnographic methodology for engagin with the Linux community. Firstly, 
we present the participant observation research method explored by the author. Then, 
we present the methodolody of the experiment done with the students.

@@

Given the tool evaluations that demonstrate the tool capabilites of finding duplications, 
we proceeded to undertand if the duplications found are possible to be mitigate and
the kernel subsystem's maintainers opinions about the duplications and their respective mitigations.
To accomplish this objective, we initialy proposed a participant-observation experiment, 
where we, people that have never contribute to the Linux kernel, 
act as contributors to the AMD Display driver and collect artifacts of our experience in the process.

As a first step of this experiment, we executed the tool in the AMD Display driver codebase and 
manually analysed the biggest duplications in terms of number of lines and raised to 
two duplicated function pairs that we judged interesting to try to mitigate. 
We have choose to approach the biggest duplications as a proof of concept that 
we can make significant impacts. The two duplicated function pairs chosen are:

@@

Given the two duplicated function pairs, we observed that the files that the functions exist contains 
multiple other duplications. Thus, we proposed a simple systematic approach to mitigate all the 
functions duplicated in the context and not only the duplicated function pairs. 
The systematic approach is defined below.

After we approached the duplications with the systematic approach. For the duplicated function
pairs that had successfully being mitigate the duplication, we sent patches to the AMD Display driver
maillist to validate the mitigations and see the maintainers opinion, while documentaion the process 
and our impressions.

@@

For each duplicated function pair, for each function in the pair, we used the \textit{function information} functionality 
(described in Section \ref{subsec:functioncommand}) to locate all duplications of the function in the codebase, 
as it may have been duplicated more times then the two occurences of the duplicated function pair. 
This strategy resulted in a collection of duplicated functions and corresponding code files.

@@

Posterior of the initial participant observation that we made, we decided to expand the experiment 
by proposing to students mitigate duplication found by the tool and send patches to the kernel community.

In the Free Software Development course at University of São Paulo, ministred by the advisor 
of this work at the first semester of 2025, we have proposed to the students the possibility
of mitigating duplications in the IIO subsystem and the AMD display driver found with the ArKanjo tool.
The course had a total of 26 students, and of that, X students being undergraduate students
and Y begin graduate students.

On the course, the students are asked to form groups of one, two or three students and together 
make a contribution to a open source software. To auxiliate the students on this task, the 
professor and teaching assistants prepares multiple alternatives on how to contribute to open source projects, 
including simple contributions as updating documentation or code style fixes. In this context, 
we have proposed two alternatives for the students related to the tool.

For the first alternative, one of the teaching assistants used the ArKanjo tool in the IIO subsystem and
listed interesting duplicated functions pair that the groups can simple mitigate this specific duplication pair, 
and it is up to the students on how
to refactor the code to mitigate the duplications and pass through the process to send patches to the IIO 
maintainers to review. For the second alternative, we propose a similar experience then the 
participant observation method, where the students pass through the whole process of executing the ArKanjo tool
in the AMD display driver, analyze the duplications found, create a patch to mitigate a duplication and sent
the patch to the AMD display driver's maintainers.

From the students enrolled in the in the course, 25 students opted for the first alternative and 1 student
opted for the second alternative, forming 11 groups working on the first alternative and 1 group 
working on the second alternative. It is worth highlighting that none of the students
that choose to mitigate duplications had made a contribution to the Linux kernel before,
meaning that the patch to mitigate the duplication being the first tentative to contribute to the kernel as
newcomers.

The groups have been request to document the experience and approaches to refactor the duplications and the
experience of submiting a patch to the Linux kernel in blogs. We had analyzed these blogs to undertand 
patterns of refactor used for mitigating the duplications, the experience on send the patches and the 
opinions of the subsystem maintainers on the mitigations.

For the teaching assistant and the student that executed the ArKanjo tool to explore duplications, 
we have asked them to answer some questions related to the experience of using the tool and what could be
improved. The questions asked is the following:

>>>>

We employ a multi-method research approach, using different methodologies to achieve our 
goals to approach code duplication in some of the Linux subsystems. To validate if the 
ArKanjo tool can find code duplications, we applied two independent methods and performed 
triangulation on the results.

The first method evaluates the proposed tool using an approach from the literature, 
comparing it against the BigCloneBench dataset \citep{bigclonebench}. 
The second method is an empirical analysis of a subset of functions within the 
AMD Display driver that our tool identified as duplicates. After completing these two 
methods, we triangulated the findings to analyze the results comprehensively. 
We chose triangulation because relying solely on the literature-based method has 
limitations and may may be unreliable \citep{bigfail, litreview}. 
As a reminder, all research approaches were done using the text similarity method 
as the code duplication method of the tool.

Given the validation of the tool’s capabilities, we proceed to understand the viability 
of mitigating the code duplication found by the tool. We conducted an ethnographic study 
using a participant-observation methodology to understand its viability. In this study, 
we aim to mitigate a subset of duplicated functions by refactoring them while interacting 
with the Linux community to gather feedback on our mitigation efforts.

The ethnographic study had two phases. In the first phase, the author explored the process 
of mitigating a code duplication found in the AMD Display driver and sent it as a PATCH to 
interact with the Linux community and get their feedback. In the second phase, we proposed 
to students in the course Free Software Development at the University of São Paulo to mitigate 
the code duplications found by the tool in the IIO subsystems and the AMD Display driver and 
pass the process of sending the mitigations as PATCH.

@@

This section outlines our approach to mitigating code duplication in the Linux kernel 
and describes the ethnographic methodology for engaging with the Linux community. 
Firstly, we present the participant observation research method explored by the author. 
Then, we present the methodology of the experiment done with the students.

@@

Given the tool evaluations demonstrating the tool's capabilities of finding duplications, 
we proceeded to understand if the duplications found can be mitigated and the kernel 
subsystem's maintainers' opinions about the duplications and their respective mitigations. 
To accomplish this objective, we initially proposed a participant-observation experiment, 
where we, people who have never contributed to the Linux kernel, act as contributors to 
the AMD Display driver and collect artifacts of our experience in the process.

As a first step of this experiment, we executed the tool in the AMD Display driver codebase. 
We manually analyzed the biggest duplications regarding the number of lines and raised them 
to two duplicated function pairs that we judged interesting to try to mitigate. We have 
chosen to approach the biggest duplications as proof of concept so that we can make 
significant impacts. The two duplicated function pairs chosen are:

@@

Given the two duplicated function pairs, we observed that the files in the functions 
contain multiple other duplications. Thus, we proposed a simple systematic approach 
to mitigate all the functions duplicated in the context, not just the duplicated 
function pairs. 
After we approached the duplications with the systematic approach, for the duplicated 
function pairs that had successfully mitigated the duplication, we sent patches to the 
AMD Display driver email list to validate the mitigations and see the maintainer's opinion 
while documenting the process and our impressions.
The systematic approach is defined below.

@@

For each duplicated function pair, for each function in the pair, we used the \textit{function information} functionality 
(described in Section \ref{subsec:functioncommand}) to locate all duplications of the function in the codebase, 
as it may have been duplicated more times than the two occurrences of the duplicated function pair. 
This strategy resulted in a collection of duplicated functions and corresponding code files.

@@

Based on our initial participant observation, we decided to expand the experiment by proposing 
to students how to mitigate duplication found by the tool and send patches to the kernel community.

In the Free Software Development course at the University of São Paulo, taught by the advisor of 
this work in the first semester of 2025, we proposed to the students the possibility of mitigating 
duplications in the IIO subsystem and the AMD display driver found with the ArKanjo tool. 
The course had 26 students, of whom X were undergraduate students and Y were graduate students.

In the course, the students are asked to form groups of one, two, or three students and contribute 
to free software together. To assist the students in this task, the professor and teaching assistants 
prepare multiple alternatives for contributing to free software projects, including simple contributions 
such as updating documentation or code style fixes. In this context, we have proposed two options 
for the students related to the tool.

For the first alternative, one of the teaching assistants used the ArKanjo tool in the IIO subsystem 
and listed interesting duplicated function pairs that the groups could mitigate this specific 
duplication pair. It is up to the students to refactor the code to mitigate the duplications and 
pass through the process of sending patches to the IIO maintainers for review. For the second 
alternative, we propose a similar experience to the participant observation method, where the 
students pass through the whole process of executing the ArKanjo tool in the AMD display driver, 
analyze the duplications found, create a patch to mitigate a duplication, and send the patch to the 
AMD display driver’s maintainers.

From the students enrolled in the course, 25 students opted for the first alternative, and one 
student opted for the second alternative, forming 11 groups working on the first alternative 
and 1 group working on the second alternative. It is worth highlighting that none of the students 
who chose to mitigate duplications had contributed to the Linux kernel before, meaning that the 
patch to mitigate the duplications was the first tentative to contribute to the kernel as newcomers.

The groups have been requesting that we document the experience and approaches to refactoring the 
duplications and the experience of submitting a patch to the Linux kernel in blogs. We analyzed 
these blogs to understand the patterns of refactoring used for mitigating the duplications, the 
experience of sending the patches, and the opinions of the subsystem maintainers on the mitigations.

The teaching assistant and the student who executed the ArKanjo tool to explore duplications were 
asked to answer some questions related to their experience using the tool and what could be improved. 
The survey with the users' answers can be found in Appendix \ref{app:survey}.

In the second stage of the Free Software Development course, we propose that students contribute 
directly to the ArKanjo tool development without help from the tool creator to verify whether the 
tool can be free software to which multiple people can contribute. The students approached the tool’s 
issue, and to our knowledge, they pointed to us in the survey feedback. Three students forming two 
groups worked in direct contributions, which was not a very expressive number compared with the 
initial experiment.

### 2025-06-01 conteudo/results/participant.tex
some work
fix grammar with grammarly

For our initial participant observation, we attempted to mitigate two code duplications found by ArKanjo tool
in the AMD Display driver applying our systematic approach presented in \ref{subsubsec:systematic}. 
As presented in section \ref{sec:meteth}, we opted for applying the systematic approach in the functions 
\textit{offset\_to\_id} and \textit{phy\_id\_to\_atom} in the folders \textit{dc/gpio} and \text{dc/bios} respectively.

Hereafter, the mitigation on \textit{offset\_to\_id} will be referred to as mitigation 1, and the mitigation on 
\textit{phy\_id\_to\_atom} will be referred to as mitigation 2. Table \ref{tab:patch} summarizes the impact of the mitigations
in terms of files and lines changed, and the refactoring method used. We present our experience and learning for each mitigation 
in the next sections.

@@

On the process of the systematic approach, we found that the \textit{offset\_to\_id} function exist in 10 files, enclosing
multiple GPU architectures. The duplicated functions are not exactly equals, as there are some minimal specific logics applied
to each GPU architecture. To address this issue, we applied the extract method to break the function in smaller parts and the 
parameterize method to address the specific logics for the GPU's.

The base idea on how we would mitigate the duplications found were done, but we could not complete the refactoring on a state 
capable to submit as a patch to the driver. 
We found that the configuration files on the AMD Display driver are designed to be imported at compilation time
using \textit{\#define} macros. The configuration files design choice makes the refactoring of duplications on generic 
approachs trickier, as refactoring code that depends on these files requires significant refactor on the configuration files
design and deep knowledge of the codebase, which us as first-timer contributors do not have. Thus, we opted to not continue 
on investigating this mitigation.

@@

\subsection{OLD}

In the first function pair, completing the refactoring requires the expertise of a maintainer for the AMD Display driver. This observation is because duplicated code artifacts involve copy-pasted code that imports different configuration files. The design of the AMD Display driver configuration files necessitates importing specific configurations at compile time, which prevents a generic approach. Consequently, refactoring this pair requires changes in how the AMD Display driver manages configuration files -- a complex task that demands input from an expert or maintainer.

The refactoring of the second function pair, however, was successful. We submitted this refactoring as a patch to the AMD Display driver to obtain feedback from the maintainers. The process of sending the patch took more time than expected. We initially submitted the patch on August 9th, 2024, but we did not receive replies. As a result, we had to resend it on October 9th, 2024 and received an initial response on November 3rd, 2024.

We understood new details about the driver evolution processes in discussions with our partner from the AMD Display driver. Some of the code in the driver is shared across all operating systems that support the implemented GPUs. This fact creates a complex process within AMD to format and submit changes across the supported systems while simultaneously implementing measures to mitigate errors.

Additionally, we comprehended that AMD Display driver developers can not view duplicated code negatively, as we understand it is a good practice in software engineering. One example shared with us is that duplicated code enhances the independence of GPU driver code, allowing developers to make changes to a specific GPU without needing to test compatibility with others. This approach helps save significant time and effort.

Regarding the patch content, only minor changes were requested to align with code 
style and best practices that were initially unknown to us. More details of the 
patch can be found in Appendix \ref{app:feedback}.

>>>>

For our initial participant observation, we attempted to mitigate two code 
duplications found by the ArKanjo tool in the AMD Display driver by applying 
our systematic approach presented in \ref{subsubsec:systematic}. 
As presented in section \ref{sec:meteth},
we opted to apply the systematic approach in the functions
\textit{offset\_to\_id} and \textit{phy\_id\_to\_atom} in the folders 
\textit{dc/gpio} and \text{dc/bios}, respectively.

Hereafter, the mitigation on \textit{offset\_to\_id} will be referred to as mitigation 1, and the mitigation on 
\textit{phy\_id\_to\_atom} will be referred to as mitigation 2. 
Table \ref{tab:patch} summarizes the impact of the mitigations in terms of 
files and lines changed and the refactoring method used. In the next sections, 
we present our experience and learning for each mitigation.

@@

In the systematic approach, we found that the \textit{offset\_to\_id} function 
exists in code 10 files, enclosing multiple GPU architectures. The duplicated 
functions are not exactly equal, as some minimal specific logic is applied to 
each GPU architecture. To address this issue, we used the extract method to 
break the function into smaller parts and the parameterized method to address 
the specific logic for the GPUs.

We reached a point with a scratch on how to mitigate the duplications found, 
but we could not complete the refactoring on a state capable of submitting as 
a patch to the driver. We found that the configuration files on the AMD Display 
driver are designed to be imported at compilation time using \textit{\#define} macros. 
The configuration files design choice makes the refactoring of duplications on 
generic approaches trickier, as refactoring code that depends on these files 
requires significant refactoring on the configuration files design and deep 
knowledge of the codebase, which we as first-timer contributors, do not have. 
Thus, we opted not to continue investigating this mitigation.

@@

Using the systematic approach, we found that the \textit{phy\_id\_to\_atom}
function exists in five code files, and two other functions are duplicated 
in those five code files. All functions across the files were exactly equal, 
so there was no need to apply the refactoring methods presented in the literature. 
The refactoring work was resumed to create a generic library and fix the compilation targets.

The code files in the second mitigation do not depend on configuration files. Thus, 
we did not face the same issues during the first mitigation. Refactoring the code 
to mitigate duplications reached a point we judged was good enough to submit as a 
patch to contribute to the AMD Display driver code quality. Thus, we moved to send 
the refactoring to the driver’s maintainers as a patch, documenting our findings and experiences 
on the process.

\begin{figure}
\includegraphics[scale=0.45]{timeline_patch}
\caption{Timeline of the PATCH submission process}
\label{fig:timeline}
\end{figure}

The experience of submitting a patch was not as we expected. The process took 7 months 
and 16 days, which is significantly slower than expected. We initially submitted the 
patch on August 9th, 2024, but we did not receive replies. Thus, we had to resend it 
on October 9th, 2024. We received an initial response on November 3rd, 2024, asking for 
minimal changes; we submitted a second version with the changes request on November 16th, 2024. 
We did not receive replies at first and need to resend the patch on February 7th, 2025. 
We received a reply on February 11th, 2025, asking for additional changes. We sent a third 
version on February 24th, 2025, and the patch was accepted and integrated into the kernel 
codebase on February 25th, 2025. Figure \ref{fig:timeline} illustrates the timeline of 
the patch submission.

We discussed with our point of contact at the AMD Display driver to understand the reason 
for the slow process of submitting a patch. We learned that some of the code in the driver 
is shared across all operating systems that support the implemented GPUs. This fact creates 
a complex process within AMD to format and submit changes across the supported systems while 
simultaneously implementing measures to mitigate errors. Additionally, we comprehended that 
AMD Display driver developers can not view duplicated code negatively, as we understand it 
is a good practice in software engineering. One example shared with us is that duplicated 
code enhances the independence of GPU driver code, allowing developers to make changes to a 
specific GPU without needing to test compatibility with others. 
This approach helps save significant time and effort.

Regarding the patch changes requested for the second version, the maintainers only asked 
for minor changes to align with code style, license correct use, and best practices that 
were initially unknown to use. For the change requested on the third version, the maintainers 
asked us to move the functions created in the generic library to an existing file instead of 
creating a new one. More details of the patch can be found in Appendix \ref{app:feedback}, 
and the patch submitted in the Linux codebase can be seen here:
\url{https://lore.kernel.org/all/20250225015532.303032-1-luanicaro@usp.br/}.

The initial plan of this research was to approach and send many mitigations to collect 
many artifacts from the driver community. Still, the slow process of sending a patch 
makes this plan unviable. Thus, we opted to redirect this research to work with students 
and refactoring duplicated function pairs found by the tool directly in the IIO and the 
AMD Display driver without passing through the whole process of the systematic approach 
proposed. This alternative allowed us to collect artifacts from the students' experiences 
and findings and not be too dependent on the interaction with the kernel community. 
Another advantage is accelerating the interaction with the kernel community because we are 
not applying the systematic approach, which translates to more minor code changes.

### 2025-06-01 conteudo/results/results.tex
fix grammar with grammarly

In this chapter we present and analyze the results found in the experiments and 
methods done through our research. Following the cronology of the research, we 
start by presenting the ArKanjo tool's evaluation through both a method from the literature
and an empirical analyses. Then, we present and discuss the experience of the initial participant
observation methodolody, where we mitigate a code duplication on the AMD Display driver and sent it as a patch.
Finally, we show and discuss our discoveries on the experiment done with students at the course in University of 
São Paulo, where we propose to students to mitigate duplications found by ArKanjo tool at IIO subsystem and the
AMD Display driver and document their founds, opinions and experiences.

>>>>

In this chapter, we present and analyze the results of the experiments and methods 
used in our research. Following the chronology of the study, we start by presenting the 
evaluation of the ArKanjo tool through both a method from the literature and an empirical 
analysis. Then, we present and discuss the experience of the initial participant 
observation methodology, where we mitigate a code duplication on the AMD Display Driver 
and send it as a patch. Finally, we show and discuss our discoveries on the experiment 
done with students at the University of São Paulo, where we propose to students to mitigate 
duplications found by the ArKanjo tool at IIO subsystem and the AMD Display Driver, and 
document their founds, opinions, and experiences.
